---
title: "Effect of epidemiological variation on dynamics"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      dev=c('png','tiff'),
                      fig.path='figures/')
library(tidyverse)
library(parallel)
```

I am struggling to understand the results when there is covariation among epidemiological parameters. 
One thing that might be worth checking is whether variation (not covariation) has any effect, since I'm not sure that's been done systematically.
I will allow different degrees of variation and compare the numerical results against the analytical expectations for $R_0$ and $S_e$ to see if one or the other seems to agree more closely.

Let's start with variation in contact rate only.
According to our analytical results, the effects of variation in contact on $R_0$ and $S_e$ can be summarized as:
\begin{align}
E[R_0(c)] &= R_0(\bar{c}), \\
E[S_e(c)] &= S_e(\hat{c}) \left(1 + \frac{1}{\hat{c}^2} \sigma_c^2\right).
\end{align}
Note that with our choices of variation ($\sigma_c = \hat{c}/2$, $\sigma_c=\hat{c}$, and $\sigma_c = 2 \hat{c}$), the expected susceptible population size will be 1.25, 2, and 5 times as large as in the no variation case. 

Comparing against the numerical results.
```{r, fig.height=5, fig.width=5, units='in', fig.cap="Effect of variation in contact rate on the dynamics of susceptible, infected, and recovered hosts. By column, we move from low to moderate to high variation. The dashed black line shows the deterministic expectation.", echo=FALSE}
## Stochastic simulation code
source("GEM_SIR_c_tau_alpha_variation.R")

## Deterministic simulation code
library(deSolve)
GEM_SIR <- function(t, y, params) {
  S <- y["S"]
  I <- y["I"]
  R <- y["R"]
  
  c <- params["c"]
  shed <- params["shed"]
  h <- params["h"]
  alpha <- params["alpha"]
  gamma <- params["gamma"]
  d <- params["d"]
  b <- params["b"]
  bs <- params["bs"]
  
  N <- S + I + R
  dS <- (b - bs*N)*N - d*S - c*shed/(h+shed)*S*I
  dI <- c*shed/(h+shed)*S*I - (d+alpha+gamma)*I
  dR <- gamma*I - d*R
  
  return(list(c(dS,dI,dR)))
}

baselineparams = c(c=.035, shed=.05, h=.15, alpha=.15, gamma=.15, d=.2, 
                   sd_c=.035, sd_s=.05, sd_a=.15, sd_g=.15, b=2.5, bs=.01)
lowvarparams <- c(c=.035, shed=.05, h=.15, alpha=.15, gamma=.15, d=.2, 
                   sd_c=.035/2, sd_s=.05/2, sd_a=.15/2, sd_g=.15/2, b=2.5, bs=.01)
highvarparams <- c(c=.035, shed=.05, h=.15, alpha=.15, gamma=.15, d=.2, 
                   sd_c=.035*2, sd_s=.05*2, sd_a=.15*2, sd_g=.15*2, b=2.5, bs=.01)

## Start the simulation with the system very near the disease free equilibrium (-2 infected individuals)
y0 <- c(S=unname((baselineparams["b"]-baselineparams["d"])/baselineparams["bs"])-2,
        I=2,
        R=0)
tmax <- 100

## Deterministic simulations
det_out <- ode(y0, times=0:100, func=GEM_SIR, parms=lowvarparams)

## Stochastic simulation seeds
set.seed(1242)
seeds <- floor(runif(100,1,1e5))

## variation in contact rate
if(!file.exists("contact_rate_variation.RDS")) {
  mclapply(seeds, 
           function(s) gillespie.SIR.var_c(tmax, lowvarparams, y0, s),
           mc.cores=10) -> out_low_var_c
  mclapply(seeds, 
           function(s) gillespie.SIR.var_c(tmax, baselineparams, y0, s),
           mc.cores=10) -> out_med_var_c
  mclapply(seeds, 
           function(s) gillespie.SIR.var_c(tmax, highvarparams, y0, s),
           mc.cores=10) -> out_high_var_c
  contact_rate_variation <- list(out_low_var_c,
                                 out_med_var_c,
                                 out_high_var_c)
  saveRDS(contact_rate_variation, file="contact_rate_variation.RDS")
} else {
  contact_rate_variation <- readRDS("contact_rate_variation.RDS")
  out_low_var_c <- contact_rate_variation[[1]]
  out_med_var_c <- contact_rate_variation[[2]]
  out_high_var_c <- contact_rate_variation[[3]]
}
  
## make some plots
par(mfrow=c(3,3), mar=c(2,2,0.5,0.5), oma=rep(0.5,4))
plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_low_var_c[[i]][,c(1,2)], col=gray(0.5))
lines(0:100, lapply(out_low_var_c, function(l) l[,2]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,2)], lwd=2, lty=2)
legend(x='topright', "S", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_med_var_c[[i]][,c(1,2)], col=gray(0.5))
lines(0:100, lapply(out_med_var_c, function(l) l[,2]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,2)], lwd=2, lty=2)
legend(x='topright', "S", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_high_var_c[[i]][,c(1,2)], col=gray(0.5))
lines(0:100, lapply(out_high_var_c, function(l) l[,2]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,2)], lwd=2, lty=2)
legend(x='topright', "S", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_low_var_c[[i]][,c(1,3)], col=gray(0.5))
lines(0:100, lapply(out_low_var_c, function(l) l[,3]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,3)], lwd=2, lty=2)
legend(x='topright', "I", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_med_var_c[[i]][,c(1,3)], col=gray(0.5))
lines(0:100, lapply(out_med_var_c, function(l) l[,3]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,3)], lwd=2, lty=2)
legend(x='topright', "I", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_high_var_c[[i]][,c(1,3)], col=gray(0.5))
lines(0:100, lapply(out_high_var_c, function(l) l[,3]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,3)], lwd=2, lty=2)
legend(x='topright', "I", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_low_var_c[[i]][,c(1,4)], col=gray(0.5))
lines(0:100, lapply(out_low_var_c, function(l) l[,4]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,4)], lwd=2, lty=2)
legend(x='topright', "R", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_med_var_c[[i]][,c(1,4)], col=gray(0.5))
lines(0:100, lapply(out_med_var_c, function(l) l[,4]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,4)], lwd=2, lty=2)
legend(x='topright', "R", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_high_var_c[[i]][,c(1,4)], col=gray(0.5))
lines(0:100, lapply(out_high_var_c, function(l) l[,4]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,4)], lwd=2, lty=2)
legend(x='topright', "R", bty='n')
```

According to our analytical results, the effects of variation in shedding on $R_0$ and $S_e$ can be summarized as:
\begin{align}
E[R_0(s)] &= R_0\left(\bar{s}\right)\left(1-\frac{h}{\bar{s}\left(h+\bar{s}\right)^2}\sigma_s^2\right), \\
E[S_e(s)] &= S_e(\hat{s}) \left(1 + \frac{h}{\hat{s}^2(h+\hat{s})} \sigma_s^2\right).
\end{align}
Based on this, we expect variation in shedding to decrease $R_0$, and to increase $S_e$.

```{r, fig.height=5, fig.width=5, units='in', fig.cap="Effect of variation in shedding rate on the dynamics of susceptible, infected, and recovered hosts. By column, we move from low to moderate to high variation. The dashed black line shows the deterministic expectation.", echo=FALSE}

## variation in shedding
if(!file.exists("shedding_rate_variation.RDS")) {
  mclapply(seeds, 
         function(s) gillespie.SIR.var_tau(tmax, lowvarparams, y0, s),
         mc.cores=10) -> out_low_var_tau
  mclapply(seeds, 
           function(s) gillespie.SIR.var_tau(tmax, baselineparams, y0, s),
           mc.cores=10) -> out_med_var_tau
  mclapply(seeds, 
           function(s) gillespie.SIR.var_tau(tmax, highvarparams, y0, s),
           mc.cores=10) -> out_high_var_tau
  shedding_rate_variation <- list(out_low_var_c,
                                 out_med_var_c,
                                 out_high_var_c)
  saveRDS(shedding_rate_variation, file="shedding_rate_variation.RDS")
} else {
  shedding_rate_variation <- readRDS("shedding_rate_variation.RDS")
  out_low_var_tau <- shedding_rate_variation[[1]]
  out_med_var_tau <- shedding_rate_variation[[2]]
  out_high_var_tau <- shedding_rate_variation[[3]]
}


par(mfrow=c(3,3), mar=c(2,2,0.5,0.5), oma=rep(0.5,4))
plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_low_var_tau[[i]][,c(1,2)], col=gray(0.5))
lines(0:100, lapply(out_low_var_tau, function(l) l[,2]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,2)], lwd=2, lty=2)
legend(x='topright', "S", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_med_var_tau[[i]][,c(1,2)], col=gray(0.5))
lines(0:100, lapply(out_med_var_tau, function(l) l[,2]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,2)], lwd=2, lty=2)
legend(x='topright', "S", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_high_var_tau[[i]][,c(1,2)], col=gray(0.5))
lines(0:100, lapply(out_high_var_tau, function(l) l[,2]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,2)], lwd=2, lty=2)
legend(x='topright', "S", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_low_var_tau[[i]][,c(1,3)], col=gray(0.5))
lines(0:100, lapply(out_low_var_tau, function(l) l[,3]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,3)], lwd=2, lty=2)
legend(x='topright', "I", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_med_var_tau[[i]][,c(1,3)], col=gray(0.5))
lines(0:100, lapply(out_med_var_tau, function(l) l[,3]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,3)], lwd=2, lty=2)
legend(x='topright', "I", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_high_var_tau[[i]][,c(1,3)], col=gray(0.5))
lines(0:100, lapply(out_high_var_tau, function(l) l[,3]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,3)], lwd=2, lty=2)
legend(x='topright', "I", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_low_var_tau[[i]][,c(1,4)], col=gray(0.5))
lines(0:100, lapply(out_low_var_tau, function(l) l[,4]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,4)], lwd=2, lty=2)
legend(x='topright', "R", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_med_var_tau[[i]][,c(1,4)], col=gray(0.5))
lines(0:100, lapply(out_med_var_tau, function(l) l[,4]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,4)], lwd=2, lty=2)
legend(x='topright', "R", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_high_var_tau[[i]][,c(1,4)], col=gray(0.5))
lines(0:100, lapply(out_high_var_tau, function(l) l[,4]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,4)], lwd=2, lty=2)
legend(x='topright', "R", bty='n')

```

According to our analytical results, the effects of variation in virulence on $R_0$ and $S_e$ can be summarized as:
\begin{align}
E[R_0(\alpha)] &= R_0\left(\bar{\alpha}\right)\left(1 + \frac{\sigma_\alpha^2}{\left(\bar{\alpha}+\gamma+d\right)^2}\right), \\
E[S_e(\alpha)] &= S_e(\hat{\alpha}).
\end{align}
Based on this, variation in virulence is expected to increase $R_0$ but have no effect on the number of susceptible hosts at equilibrium.


```{r, fig.height=5, fig.width=5, units='in', fig.cap="Effect of variation in virulence on the dynamics of susceptible, infected, and recovered hosts. By column, we move from low to moderate to high variation. The dashed black line shows the deterministic expectation.", echo=FALSE}
## variation in virulence
if(!file.exists("virulence_variation.RDS")) {
  mclapply(seeds, 
           function(s) gillespie.SIR.var_alpha(tmax, lowvarparams, y0, s),
           mc.cores=10) -> out_low_var_alpha
  mclapply(seeds, 
           function(s) gillespie.SIR.var_alpha(tmax, baselineparams, y0, s),
           mc.cores=10) -> out_med_var_alpha
  mclapply(seeds, 
           function(s) gillespie.SIR.var_alpha(tmax, highvarparams, y0, s),
           mc.cores=10) -> out_high_var_alpha
  virulence_variation <- list(out_low_var_alpha,
                              out_med_var_alpha,
                              out_high_var_alpha)
  saveRDS(virulence_variation, file="virulence_variation.RDS")
} else {
  virulence_variation <- readRDS("virulence_variation.RDS")
  out_low_var_alpha <- virulence_variation[[1]]
  out_med_var_alpha <- virulence_variation[[2]]
  out_high_var_alpha <- virulence_variation[[3]]
}

par(mfrow=c(3,3), mar=c(2,2,0.5,0.5), oma=rep(0.5,4))
plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_low_var_alpha[[i]][,c(1,2)], col=gray(0.5))
lines(0:100, lapply(out_low_var_alpha, function(l) l[,2]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,2)], lwd=2, lty=2)
legend(x='topright', "S", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_med_var_alpha[[i]][,c(1,2)], col=gray(0.5))
lines(0:100, lapply(out_med_var_alpha, function(l) l[,2]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,2)], lwd=2, lty=2)
legend(x='topright', "S", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_high_var_alpha[[i]][,c(1,2)], col=gray(0.5))
lines(0:100, lapply(out_high_var_alpha, function(l) l[,2]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,2)], lwd=2, lty=2)
legend(x='topright', "S", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_low_var_alpha[[i]][,c(1,3)], col=gray(0.5))
lines(0:100, lapply(out_low_var_alpha, function(l) l[,3]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,3)], lwd=2, lty=2)
legend(x='topright', "I", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_med_var_alpha[[i]][,c(1,3)], col=gray(0.5))
lines(0:100, lapply(out_med_var_alpha, function(l) l[,3]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,3)], lwd=2, lty=2)
legend(x='topright', "I", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_high_var_alpha[[i]][,c(1,3)], col=gray(0.5))
lines(0:100, lapply(out_high_var_alpha, function(l) l[,3]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,3)], lwd=2, lty=2)
legend(x='topright', "I", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_low_var_alpha[[i]][,c(1,4)], col=gray(0.5))
lines(0:100, lapply(out_low_var_alpha, function(l) l[,4]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,4)], lwd=2, lty=2)
legend(x='topright', "R", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_med_var_alpha[[i]][,c(1,4)], col=gray(0.5))
lines(0:100, lapply(out_med_var_alpha, function(l) l[,4]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,4)], lwd=2, lty=2)
legend(x='topright', "R", bty='n')

plot.new()
plot.window(xlim=c(0,100), ylim=c(0,230))
box('plot'); axis(1); axis(2)
for (i in 1:100) lines(out_high_var_alpha[[i]][,c(1,4)], col=gray(0.5))
lines(0:100, lapply(out_high_var_alpha, function(l) l[,4]) %>% do.call("cbind.data.frame",.) %>% apply(., 1, function(r) mean(r,na.rm=TRUE)), lwd=2, col=2)
lines(det_out[,c(1,4)], lwd=2, lty=2)
legend(x='topright', "R", bty='n')

```
         


